---
title: "Data import and Wrangling"
output: github_document
Date: '`r format(Sys.time(), "%Y-%m-%d")`'
name: Wen Dai
---

```{r}
library (tidyverse)
library(dplyr)
```

# Step 1: Data import.
1) In this step I download and renamed the Mental_Health_Care_in_the_Last_4_Weeks.csv to mental_health.csv to my computer, then I imported it to R and name the dataset as rawdata.
2) during this step, I removing Unnecessary Columns including : phase, quartile range and suppression flag. Then, I also renamed some columns: Improving column names for clarity or consistency.

```{r}
rawdata = read_csv("data/mental_health.csv") |> 
  janitor::clean_names() |>
  mutate(indicator = str_replace(indicator, ", Last 4 Weeks", "")) |> 
  select(-phase,-quartile_range,-suppression_flag) |> 
  mutate(group = str_replace(group, "By", ""))|> 
  select(start_date=time_period_start_date,end_date=time_period_end_date,everything()) 

```

# Step 2: create a tidy date. 
1) Create a new column named year and convert it into numeric data.
2) Convert dates to date format and change separate the time_period_label into two columns: month_period_label and years. Then delete years (because this years variable is unable convert into numeric without causing missing value problem, so I have to delete it)
3) make a comment.

```{r}

tidydata=rawdata |> 
  mutate(start_dates=mdy(pull(rawdata, start_date)),
         end_dates=mdy(pull(rawdata, end_date)))|> 
        mutate(tpl=time_period_label) |>  
        mutate(year = as.numeric(str_extract(tpl, "\\d{4}")))|> 
         separate(col=time_period_label, into=c("month_period_label","years"), sep = ", ", remove = FALSE,extra="merge") |> 
  select(-time_period_label,-start_date,-end_date) |>
  mutate(week_number=time_period-12) |> select(indicator,year,start_dates,end_dates,week_number,month_period_label,group,state,subgroup,value,low_ci,high_ci,confidence_interval)

```

### Comment:

There are 13 columns in the rawdata:

* Indicator (chr): The type of mental health care.
* year(num): year for the time period.
* Start Date (Date): Start date of the time period.
* End Date(Date): End date of the time period.
* Week_number(num):Numeric code for the time period minus 12. 
* Month Period Label(chr): Label for the month time period.
* Group(chr): The demographic group.
* State(chr): The state or national level data.
* Subgroup(chr): The specific subgroup within the group.
* Value(num): The value (percentage or number) representing the indicator.
* LowCI(num): Lower bound of the confidence interval.
* HighCI(num): Upper bound of the confidence interval.
* Confidence Interval(chr): Confidence interval as a range.
