---
title: "EDA"
author: "Sarah Younes"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message = FALSE, echo = FALSE}
library(tidyverse)

library(viridisLite)
viridisLite::viridis(10)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

```{r message = FALSE, echo = FALSE}
tidydata = 
  read.csv("data/tidydata.csv") |>
  mutate(
    start_dates = as.Date(start_dates))
```

### Gender

Trends in mental health access by gender were stark. Women reported needing almost twice as much counseling or therapy for mental health as men, and these trends were consistent from 2020 to 2022.

Women took more prescription medication for mental health and received more counseling or therapy for mental health than men. These trends slightly increased each year from 2020 to 2022.

```{r gender, echo = FALSE}
tidydata |>
  filter(
    group == "Sex",
    state == "United States") |>
  group_by(indicator, start_dates, subgroup) |>
  ggplot(aes(x = start_dates, y = value, color = indicator)) +
  geom_line() +
  facet_grid(. ~ subgroup) +
  theme_gray() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  guides(color = guide_legend(nrow = 2)) +
    labs(
    x = "Date",
    y = "Prevalence (%)",
    title = "Differences in Mental Health Access by Gender",
    caption = "Data was collected from August 19, 2020 to May 9, 2022")
```

### Sexual Orientation

Mental health access strongly differed by sexual orientation. Bisexual Americans reported the highest trends in both taking prescription medication and/or receiving counseling for mental health and needing counseling or therapy but not receiving it. Their outcomes showed the most variation over time. Straight Americans consistently reported the least unmet need for counseling or therapy.

```{r sexuality, echo = FALSE}
tidydata |>
    filter(group == "Sexual orientation",
         state == "United States",
         week_number %in% c(24:33)) |>
  group_by(start_dates, subgroup, indicator) |>
  ggplot(aes(x = start_dates, y = value, color = indicator)) +
  geom_line() +
  facet_grid(. ~ subgroup) +
  theme_grey() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  guides(color = guide_legend(nrow = 2)) +
    labs(
    x = "Date",
    y = "Prevalence (%)",
    title = "Differences in Mental Health Access by Sexual Orientation",
    caption = "Data was collected from August 18, 2021 to May 9, 2022")
```

### Race/ethnicity

The graph below explores differences in outcome variables by race/ethnicity. Trends by racial and ethnic groups were stark across all three years. Non-Hispanic Multiracial Americans, closely followed by non-Hispanic White Americans, received the most prescription medication and/or counseling or therapy for mental health. However, non-Hispanic multiracial Americans needed counseling or therapy for mental health but did not receive it.

```{r race ethnicity, echo = FALSE}
tidydata |>
  filter(group == "Race/Hispanic ethnicity", state == "United States") |>
  mutate(
    subgroup = replace(subgroup, subgroup == "Non-Hispanic Asian, single race", "Non-Hispanic Asian"),
    subgroup = replace(subgroup, subgroup == "Non-Hispanic Black, single race", "Non-Hispanic Black"),
    subgroup = replace(subgroup, subgroup == "Non-Hispanic White, single race", "Non-Hispanic White"),
    subgroup = replace(subgroup, subgroup == "Non-Hispanic, other races and multiple races", "Non-Hispanic Multi-racial")) |>
  group_by(indicator, start_dates, subgroup) |>
  ggplot(aes(x = start_dates, y = value, color = indicator)) +
  geom_line() +
  facet_grid(. ~ subgroup) +
  theme_grey() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  guides(color = guide_legend(nrow = 2)) +
  labs(
    x = "Date",
    y = "Prevalence (%)",
    title = "Differences in Mental Health Access by Race",
    caption = "Data was collected from August 19, 2020 to May 9, 2022")
```