---
title: "Exploratory Data Analysis (EDA)"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: "hide"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message = FALSE, echo = FALSE}
library(tidyverse)
library(sf)
library(maps)
library(plotly)

library(viridisLite)
viridisLite::viridis(10)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

tidydata = 
  read.csv("data/tidydata.csv") |>
  mutate(
    start_dates = as.Date(start_dates))

## visualizations of US map require install of sf and maps packages
```

## National Averages

```{r}
avg_year =
  tidydata |> 
  filter(state == "United States") |> 
  select(indicator, year, week_number, value, start_dates) |> 
  group_by(year, start_dates, indicator) |> 
  summarize(
    mean = mean(value),
    indicator_total = n())

# plot average for each start date for each indicator
avg_year |> 
  ggplot(aes(x = start_dates, y = mean, color = indicator)) +
  geom_point() +
  labs(
    x = "Start Date",
    y = "Average Value",
    title = "National Average (2020-2022)") +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle=90, hjust=1),
    strip.text = element_text(size = 4)) +
  guides(color = guide_legend(nrow = 2))
```

On average, those who took prescription medication for mental health and/or received counseling or therapy had a highest values while those who received counseling or therapy had the lowest value. 

## Confidence Intervals

I relabeled the observations for the variable indicator to "received" if they did obtain the mental health services they needed and "not received" if they did not obtain the mental health services they needed.
```{r}
received_service = 
  data |> 
  mutate(
    resolution = case_when(
      indicator == "Took Prescription Medication for Mental Health" ~ "received",
      indicator == "Received Counseling or Therapy" ~ "received",
      indicator == "Took Prescription Medication for Mental Health And/Or Received Counseling or Therapy" ~ "received",
      indicator == "Needed Counseling or Therapy But Did Not Get It" ~ "not received"))
```

I created a data frame where I grouped by states to determine the number of total services, which includes the services received as well as the number of services not received.
```{r}
states_df =
  received_service |> 
  select(state, indicator, resolution) |> 
  filter(state != "United States") |> 
  group_by(state) |> 
  summarize(
    services_total = n(),
    services_not_received = sum(resolution == "not received"))
```

Now, I focus on the state of New York. Using the `prop.test` and `broom::tidy` functions, I obtain an estimate and CI of the proportion of mental health services not received in New York (shown in the table below).

```{r}
ny_test = 
  prop.test(
    x = filter(states_df, state == "New York") %>% pull(services_not_received),
    n = filter(states_df, state == "New York") %>% pull(services_total)) 

broom::tidy(ny_test) %>% 
  knitr::kable(digits = 3)
```


I apply prop.test and broom:tidy to obtain estimates and confidence intervals for the proportion of mental health services not received for each state. 
```{r}
services_states =
  states_df |> 
  mutate(
    prop_tests = map2(services_not_received, services_total, \(x, y) prop.test(x = x, n = y)),
    tidy_tests = map(prop_tests, broom::tidy)) |> 
  select(-prop_tests) |> 
  unnest(tidy_tests) |> 
  select(state, estimate, conf.low, conf.high) |> 
  mutate(state = fct_reorder(state, estimate))
```


Lastly, below is a plot showing the estimate (and CI) of the proportion of mental health services not received in each state.

```{r}
services_states %>% 
  mutate(state = fct_reorder(state, estimate)) %>% 
  ggplot(aes(x = state, y = estimate)) + 
  geom_point() + 
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high)) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```


## Disaggregations by State

This EDA will analyze mental health care access trends by state.

I will first create a dataframe that only includes state-level observations.

```{r}
state_df =
  tidydata |> 
  filter(group == "State")
```

### Indicator 1: Took prescription

Let's start with the mental health indicator "Took Prescription Medication for Mental Health"

The following chunk shows the top five states with the highest mean percentage of respondents who reported taking prescription medication for mental health from 2020-2022.

```{r}
state_meds_top5 =
  state_df |> 
  filter(indicator == "Took Prescription Medication for Mental Health") |> 
  group_by(state) |> 
  summarize(mean_value = mean(value)) |> 
  arrange(desc(mean_value)) |> 
  top_n(5) 

state_meds_top5 |> 
  knitr::kable()
```

Let's visualize a heat map of this indicator's prevalence by US state.

First, we'll merge the filtered dataset with an existing US state map dataset to assist with mapping.

```{r}
state_meds =
  state_df |> 
  filter(indicator == "Took Prescription Medication for Mental Health") |> 
  group_by(state) |> 
  summarize(mean_value = mean(value)) |> 
  mutate(state = tolower(state))

usa = 
  st_as_sf(map("state", fill = TRUE, plot = FALSE))

us_meds =
  merge(usa, state_meds, by.x = "ID", by.y = "state", all.x = TRUE)
```

Next, we'll plot the data using ggplot

```{r}
heatmap_meds=
ggplot(us_meds)+
  geom_sf(aes(fill = mean_value), color = "white", size = 0.2)+
  scale_fill_gradient(low = "white", high = "red", na.value = "grey50", name = "Avg % taking prescription medications")+
  theme_minimal()+
  theme(legend.position = "right",
        plot.title.position = "plot")+
  labs(title = "Average % used prescription medication for mental health 2020-2022, by state")

print(heatmap_meds)
```

We can also look at which states have had the greatest change in medication use. This shows that not only does WV have the highest prevalence of medication use for mental health during this period, but it also has had the highest level of change in this indicator between 2020-2022.

```{r}
state_meds_change =
  state_df |> 
  filter(indicator == "Took Prescription Medication for Mental Health", year != "2021") |> 
  group_by(state, year) |> 
  summarize(mean_value = mean(value)) |> 
  arrange(state, year) |> 
  mutate(val_change = c(NA, diff(mean_value)))

state_meds_change |> 
  filter(!is.na(val_change)) |> 
  ungroup() |> 
  select(-year, -mean_value) |> 
  slice_max(order_by = val_change, n=5) |> 
  knitr::kable()
  
```

Let's take a look at the trend for this indicator over time. 

```{r, fig.keep="none"}

state_df |>
  filter(indicator == "Took Prescription Medication for Mental Health") |> 
  group_by(state, year) |>
  summarize(mean_value = mean(value)) |> 
  mutate(year = factor(year)) |> 
plot_ly(x = ~year, y = ~mean_value, color = ~state, type = "scatter", mode = "lines+markers") |>
  layout(
    title = "% used prescription medication for mental health by state & year",
    xaxis = list(title = "Year"),
    yaxis = list(title = "Average % used prescription medication for mental health"))

```

### Indicator 2: Received Counseling or Therapy

Now we'll look at the mental health indicator "Received Counseling or Therapy"

The following chunk shows the top five states with the highest mean percentage of respondents who reported receiving counseling or therapy from 2020-2022. The state with the highest percentage during this period was District of Columbia.

```{r}

state_ther_top5 =
  state_df |> 
  filter(indicator == "Received Counseling or Therapy") |> 
  group_by(state) |> 
  summarize(mean_value = mean(value)) |> 
  arrange(desc(mean_value)) |> 
  top_n(5) 

state_ther_top5 |> 
  knitr::kable()

```

Let's visualize a heat map of this indicator's prevalence by US state.

First, we'll merge the filtered dataset with an existing US state map dataset to assist with mapping.

```{r}
state_ther =
  state_df |> 
  filter(indicator == "Received Counseling or Therapy") |> 
  group_by(state) |> 
  summarize(mean_value = mean(value)) |> 
  mutate(state = tolower(state))

usa = 
  st_as_sf(map("state", fill = TRUE, plot = FALSE))

us_ther =
  merge(usa, state_ther, by.x = "ID", by.y = "state", all.x = TRUE)
```

Next, we'll plot the data using ggplot

```{r}
heatmap_ther=
ggplot(us_ther)+
  geom_sf(aes(fill = mean_value), color = "white", size = 0.2)+
  scale_fill_gradient(low = "white", high = "red", na.value = "grey50", name = "Avg % receiving therapy")+
  theme_minimal()+
  theme(legend.position = "right",
        plot.title.position = "plot")+
  labs(title = "Average % receiving counseling or therapy 2020-2022, by state")

print(heatmap_ther)
```

We can also look at which states have had the greatest change in therapy use. This shows that NE had the highest level of change in this indicator between 2020-2022. It also had a high level of change in indicator 1

```{r}
state_ther_change =
  state_df |> 
  filter(indicator == "Received Counseling or Therapy", year != "2021") |> 
  group_by(state, year) |> 
  summarize(mean_value = mean(value)) |> 
  arrange(state, year) |> 
  mutate(val_change = c(NA, diff(mean_value)))

state_ther_change |> 
  filter(!is.na(val_change)) |> 
  ungroup() |> 
  select(-year, -mean_value) |> 
  slice_max(order_by = val_change, n=5) |> 
  knitr::kable()
```

Let's take a look at the trend for this indicator over time. 

```{r, fig.keep="none"}
state_df |>
  filter(indicator == "Received Counseling or Therapy") |> 
  group_by(state, year) |>
  summarize(mean_value = mean(value)) |> 
  mutate(year = factor(year)) |> 
plot_ly(x = ~year, y = ~mean_value, color = ~state, type = "scatter", mode = "lines+markers") |>
  layout(
    title = "% received counseling or therapy by state & year",
    xaxis = list(title = "Year"),
    yaxis = list(title = "Average % received therapy"))
```

### Indicator 3: "Took Prescription Medication for Mental Health And/Or Received Counseling or Therapy"

The third indicator we'll look at combines the first two indicators into one category. This can help give a snapshot of overall mental health care need.

The following chunk shows the top five states with the highest mean percentage of respondents who reported mental health medication and/or therapy use from 2020-2022. The state with the highest percentage during this period was West Virginia.

```{r}
state_both_top5 =
  state_df |> 
  filter(indicator == "Took Prescription Medication for Mental Health And/Or Received Counseling or Therapy") |> 
  group_by(state) |> 
  summarize(mean_value = mean(value)) |> 
  arrange(desc(mean_value)) |> 
  top_n(5) 

state_both_top5 |> 
  knitr::kable()
```

Let's visualize a heat map of this indicator's prevalence by US state.

First, we'll merge the filtered dataset with an existing US state map dataset to assist with mapping.

```{r}
state_both =
  state_df |> 
  filter(indicator == "Took Prescription Medication for Mental Health And/Or Received Counseling or Therapy") |> 
  group_by(state) |> 
  summarize(mean_value = mean(value)) |> 
  mutate(state = tolower(state))

usa = 
  st_as_sf(map("state", fill = TRUE, plot = FALSE))

us_both =
  merge(usa, state_both, by.x = "ID", by.y = "state", all.x = TRUE)
```

Next, we'll plot the data using ggplot

```{r}
heatmap_both=
ggplot(us_both)+
  geom_sf(aes(fill = mean_value), color = "white", size = 0.2)+
  scale_fill_gradient(low = "white", high = "red", na.value = "grey50", name = "Avg % medication and/or therapy use")+
  theme_minimal()+
  theme(legend.position = "right",
        plot.title.position = "plot")+
  labs(title = "Average % using medication and/or therapy 2020-2022, by state")

print(heatmap_both)
```

We can also look at which states have had the greatest change in mental health care use. This shows that WV had the highest level of change in this indicator between 2020-2022. This makes sense, as WV had the highest change in medication use.

```{r}
state_both_change =
  state_df |> 
  filter(indicator == "Took Prescription Medication for Mental Health And/Or Received Counseling or Therapy", year != "2021") |> 
  group_by(state, year) |> 
  summarize(mean_value = mean(value)) |> 
  arrange(state, year) |> 
  mutate(val_change = c(NA, diff(mean_value)))

state_both_change |> 
  filter(!is.na(val_change)) |> 
  ungroup() |> 
  select(-year, -mean_value) |> 
  slice_max(order_by = val_change, n=5) |> 
  knitr::kable()
```

Let's take a look at the trend for this indicator over time. 

```{r, fig.keep="none"}
state_df |>
  filter(indicator == "Took Prescription Medication for Mental Health And/Or Received Counseling or Therapy") |> 
  group_by(state, year) |>
  summarize(mean_value = mean(value)) |> 
  mutate(year = factor(year)) |> 
plot_ly(x = ~year, y = ~mean_value, color = ~state, type = "scatter", mode = "lines+markers") |>
  layout(
    title = "% medication and/or therapy use by state & year",
    xaxis = list(title = "Year"),
    yaxis = list(title = "Average % medication and/or therapy use"))
```

### Indicator 4: Needed Counseling or Therapy But Did Not Get It

The last indicator we'll look at is describes gaps in mental health care access, specifically counseling and therapy. 

The following chunk shows the top five states with the highest mean percentage of respondents who reported no receipt of needed counseling/therapy from 2020-2022. The state with the highest percentage during this period was Oregon. Surprisingly, District of Columbia also appeared on the top 5, even though it was the "state" with the highest level of therapy use during this period.

```{r}
state_need_top5 =
  state_df |> 
  filter(indicator == "Needed Counseling or Therapy But Did Not Get It") |> 
  group_by(state) |> 
  summarize(mean_value = mean(value)) |> 
  arrange(desc(mean_value)) |> 
  top_n(5) 

state_need_top5 |> 
  knitr::kable()
```

Let's visualize a heat map of this indicator's prevalence by US state.

First, we'll merge the filtered dataset with an existing US state map dataset to assist with mapping.

```{r}
state_need =
  state_df |> 
  filter(indicator == "Needed Counseling or Therapy But Did Not Get It") |> 
  group_by(state) |> 
  summarize(mean_value = mean(value)) |> 
  mutate(state = tolower(state))

usa = 
  st_as_sf(map("state", fill = TRUE, plot = FALSE))

us_need =
  merge(usa, state_need, by.x = "ID", by.y = "state", all.x = TRUE)
```

Next, we'll plot the data using ggplot

```{r}
heatmap_need=
ggplot(us_need)+
  geom_sf(aes(fill = mean_value), color = "white", size = 0.2)+
  scale_fill_gradient(low = "white", high = "red", na.value = "grey50", name = "Avg % did not receive needed therapy")+
  theme_minimal()+
  theme(legend.position = "right",
        plot.title.position = "plot")+
  labs(title = "Average % did not receive needed therapy 2020-2022, by state")

print(heatmap_need)
```

We can also look at which states have had the greatest change in unmet mental health care need. This shows that AK had the highest level of change in this indicator between 2020-2022.

```{r}
state_need_change =
  state_df |> 
  filter(indicator == "Needed Counseling or Therapy But Did Not Get It", year != "2021") |> 
  group_by(state, year) |> 
  summarize(mean_value = mean(value)) |> 
  arrange(state, year) |> 
  mutate(val_change = c(NA, diff(mean_value)))

state_need_change |> 
  filter(!is.na(val_change)) |> 
  ungroup() |> 
  select(-year, -mean_value) |> 
  slice_max(order_by = val_change, n=5) |> 
  knitr::kable()
```

Let's take a look at the trend for this indicator over time. 

```{r, fig.keep="none"}
state_df |>
  filter(indicator == "Needed Counseling or Therapy But Did Not Get It") |> 
  group_by(state, year) |>
  summarize(mean_value = mean(value)) |> 
  mutate(year = factor(year)) |> 
plot_ly(x = ~year, y = ~mean_value, color = ~state, type = "scatter", mode = "lines+markers") |>
  layout(
    title = "% did not receive needed therapy by state & year",
    xaxis = list(title = "Year"),
    yaxis = list(title = "Avg % did not receive needed therapy"))
```

### Discussion

After analysis of the four indicators by state, we found the lowest prevalence of mental health care use in some of the most populous US states: California, New York, Texas, and Florida. Medication use prevalence was highest in West Virginia, Kentucky, Arkansas, Utah, and Oklahoma. Except for Utah, these states have high rates of [poverty](https://data.ers.usda.gov/reports.aspx?ID=17826) and low [median household income] (https://www.kff.org/other/state-indicator/median-annual-income/?activeTab=map&currentTimeframe=0&selectedDistributions=median-annual-household-income&sortModel=%7B%22colId%22:%22Location%22,%22sort%22:%22asc%22%7D), which could drive higher rates of mental health medication use.

Therapy use was highest in District of Columbia, Massachusetts, Rhode Island, Vermont, and Oregon. These states have high levels of [educational attainment](https://data.ers.usda.gov/reports.aspx?ID=17829#P9452d1b2fc324fd3ba7da7fb1edf4e99_2_240iT2) which could mean that their populations have more positive perceptions of therapy. That said, the greatest change in therapy use during the pandemic was in more rural states with lower median household income, like Nebraska and North Dakota. It’s possible that these were states that particularly benefited from expansions in telehealth for providing therapy. 
When medication and therapy indicators are combined to indicate overall mental health burden, West Virginia, Utah, and Kentucky remain the highest, but are joined by Vermont and Maine. Vermont, Maine, and Utah all have high median household income, which suggests ruralness rather than income levels may drive mental health burden at the state level.

Interestingly, the states with the highest prevalence of unmet therapy needs also tended to have a higher prevalence of therapy use. This could indicate that demand for therapy outpaced supply of mental health care providers even with the expansion of telehealth during the pandemic. However, West Virginia, which had a high prevalence of medication use, did not have high unmet therapy needs. This could mean that in certain states, medication use is not tied to therapy use.