---
title: "EDA_by_state"
author: "Caleigh Dwyer"
date: "2023-11-26"
output: github_document
---

```{r}
library(tidyverse)
```

import tidy dataset:

```{r}
##note: should we export tidy dataset to our directory on github so we can delete this code from any EDA?

rawdata =
  read_csv("data/mental_health.csv")

tidydata = 
  rawdata |> 
  janitor::clean_names() |>
  mutate(indicator = str_replace(indicator, ", Last 4 Weeks", "")) |> 
  select(-phase, -quartile_range, -suppression_flag) |> 
  mutate(group = str_replace(group, "By ", "")) |>
  drop_na(value) |>
  select(start_date = time_period_start_date, end_date = time_period_end_date, everything())

tidydata =
  tidydata |>
  mutate(week_number = match(time_period_label, unique(time_period_label))) |>
  mutate(
    start_dates = mdy(pull(tidydata, start_date)),
    end_dates = mdy(pull(tidydata, end_date))) |> 
  mutate(tpl = time_period_label) |>  
  mutate(year = as.numeric(str_extract(tpl, "\\d{4}")))|> 
  separate(col = time_period_label, into = c("week_label", "years"), sep = ", ", remove = FALSE, extra = "merge") |>
  select(indicator, year, start_dates, end_dates, week_number, week_label = time_period_label, state, group, subgroup, value, low_ci, high_ci, confidence_interval)
```



## EDA

This EDA will analyze mental health care access trends by state.

I will first create a dataframe that only includes state-level observations.

```{r}
state_df =
  tidydata |> 
  filter(group == "State")
```

### Indicator 1: Took prescription

Let's start with the mental health indicator "Took Prescription Medication for Mental Health"

The following chunk shows the top five states with the highest mean percentage of respondents who reported taking prescription medication for mental health from 2020-2022.

```{r}

state_meds =
  state_df |> 
  filter(indicator == "Took Prescription Medication for Mental Health") |> 
  group_by(subgroup) |> 
  summarize(mean_value = mean(value)) |> 
  arrange(desc(mean_value)) |> 
  top_n(5) |> 
  knitr::kable()

```




 "Took Prescription Medication for Mental Health And/Or Received Counseling or Therapy"

