---
title: "Exploratory Data Analysis (EDA)"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: "hide"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message = FALSE, echo = FALSE}
library(tidyverse)

library(viridisLite)
viridisLite::viridis(10)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

tidydata = 
  read.csv("data/tidydata.csv") |>
  mutate(
    start_dates = as.Date(start_dates))
```

## Age Group

I started by filtering the tidy dataset to only include age and residence in the United States. Then, I renamed the "80 years and above" subgroup to 80+ years. Afterward, I plotted the age group graph; however, because the start_date variable overlapped, the legend was difficult to view when I plotted the graph. As a result, I rotated the x-axis labels vertically and stacked the legend to two columns.

```{r age}
#Filtering the dataset to age and united states
age_plot <- tidydata |> 
  filter(group == "Age", state == "United States")

#Renaming the "80 years and above" subgroup to 80+ years
age_plot$subgroup[age_plot$subgroup == "80 years and above"] = "80+ years"

#Ploting the age group
age_plot |> 
  group_by(indicator, start_dates, subgroup) |>
  ggplot(aes(x = start_dates, y = value, color = indicator)) +
  geom_line() +
  facet_grid(. ~ subgroup) +
  labs(
    x = "Start Date",
    y = "Average Value",
    title = "Differences in Indicators Among Age Groups (2021-2022)") +
  theme_gray() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)  # Rotate x-axis labels vertically
  ) +
  scale_color_hue(h = c(100, 300)) +
  guides(color = guide_legend(nrow = 2)) # Adding the legend guide adjustment
```

Critical findings:

Based on the graph, it is clear that among the age groups, 18-29 have the highest values across all the indicators. Especially for the indicator of taking prescription medication for mental health and or receiving counseling, followed by the 30-39 age group, then 40-49 years. Since the data is captured from 2021 to 2022 during the pandemic, lockdown and transition to a virtual environment might cause tremendous pressure on these age groups. In addition, limited access to social life might burden the 18-29 age group. In addition, it is surprising to see that age groups from 60 and above have the lowest values of needed counseling or therapy but did not get it and received counseling or therapy. Maybe many of them are in their retirement, or this age group has a stigma around mental health because they might have a negative impression about having mental health issues. Therefore, they are less likely to admit they need counseling or therapy. 

## Education Level 

I began by filtering the tidy dataset to only age and participants residing in the United States. In the next step, I re-leveled the education subgroup by creating a new variable, ed_name. Afterward, I used the new variable, ed_name, to create an education plot. Similar to the age plot, the x-axis labels and legend are difficult to view. So, I adjusted the x-axis labels to vertical and stacked the legend columns into two columns. 

```{r education}
#Filtering the dataset to age and united states, re-level the subgroup, and plot the education plot 
ed_plot= 
  tidydata |>
  filter(group == "Education", state == "United States") |> 
   mutate(ed_name = forcats::fct_relevel(subgroup, c("Bachelor's degree or higher",
                                                     "Some college/Associate's degree", "High school diploma or GED", "Less than a high school diploma"))) |> 
  group_by(start_dates, ed_name, indicator) |> 
  ggplot(aes(x = start_dates, y = value, color = indicator)) +
  geom_line() +
  facet_grid(. ~ ed_name) +
  labs(
    x = "Start Date",
    y = "Average Value",
    title = "Differences in Indicators by Education Levels (2021-2022)") +
  theme_gray() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)  # Rotate x-axis labels vertically
  ) +
  scale_color_hue(h = c(100, 300)) +
  guides(color = guide_legend(nrow = 2)) # Adding the legend guide adjustment

#View the education plot
ed_plot
```

Critical findings:

Among four education groups, the bachelor's degree or higher and some college/ associate's degree groups have the highest value of taking prescription medication for mental health and/or received counseling. This is because individuals with higher education typically have better jobs, more money, and many other benefits, including better health insurance, which leads to better access to quality mental health services than those with a lower education level. On the other hand, those who have a lower education level (high school diploma or GED and less than high school diploma groups) have the least value of needed counseling or therapy but did not get it and received counseling or therapy indicators because they might not have adequate resources to the services which resulting having the lowest values in term of the utilization of the counseling services.

## Gender

Trends in mental health access by gender were stark. Women reported needing almost twice as much counseling or therapy for mental health as men, and these trends were consistent from 2020 to 2022.

Women took more prescription medication for mental health and received more counseling or therapy for mental health than men. These trends slightly increased each year from 2020 to 2022.

```{r gender}
tidydata |>
  filter(
    group == "Sex",
    state == "United States") |>
  group_by(indicator, start_dates, subgroup) |>
  ggplot(aes(x = start_dates, y = value, color = indicator)) +
  geom_line() +
  facet_grid(. ~ subgroup) +
  theme_gray() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  guides(color = guide_legend(nrow = 2)) +
    labs(
    x = "Date",
    y = "Prevalence (%)",
    title = "Differences in Mental Health Access by Gender",
    caption = "Data was collected from August 19, 2020 to May 9, 2022")
```

## Sexual Orientation

Mental health access strongly differed by sexual orientation. Bisexual Americans reported the highest trends in both taking prescription medication and/or receiving counseling for mental health and needing counseling or therapy but not receiving it. Their outcomes showed the most variation over time. Straight Americans consistently reported the least unmet need for counseling or therapy.

```{r sexuality}
tidydata |>
    filter(group == "Sexual orientation",
         state == "United States",
         week_number %in% c(24:33)) |>
  group_by(start_dates, subgroup, indicator) |>
  ggplot(aes(x = start_dates, y = value, color = indicator)) +
  geom_line() +
  facet_grid(. ~ subgroup) +
  theme_grey() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  guides(color = guide_legend(nrow = 2)) +
    labs(
    x = "Date",
    y = "Prevalence (%)",
    title = "Differences in Mental Health Access by Sexual Orientation",
    caption = "Data was collected from August 18, 2021 to May 9, 2022")
```

## Race/Ethnicity

The graph below explores differences in outcome variables by race/ethnicity. Trends by racial and ethnic groups were stark across all three years. Non-Hispanic Multiracial Americans, closely followed by non-Hispanic White Americans, received the most prescription medication and/or counseling or therapy for mental health. However, non-Hispanic multiracial Americans needed counseling or therapy for mental health but did not receive it.

```{r race ethnicity}
tidydata |>
  filter(group == "Race/Hispanic ethnicity", state == "United States") |>
  mutate(
    subgroup = replace(subgroup, subgroup == "Non-Hispanic Asian, single race", "Non-Hispanic Asian"),
    subgroup = replace(subgroup, subgroup == "Non-Hispanic Black, single race", "Non-Hispanic Black"),
    subgroup = replace(subgroup, subgroup == "Non-Hispanic White, single race", "Non-Hispanic White"),
    subgroup = replace(subgroup, subgroup == "Non-Hispanic, other races and multiple races", "Non-Hispanic Multi-racial")) |>
  group_by(indicator, start_dates, subgroup) |>
  ggplot(aes(x = start_dates, y = value, color = indicator)) +
  geom_line() +
  facet_grid(. ~ subgroup) +
  theme_grey() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  guides(color = guide_legend(nrow = 2)) +
  labs(
    x = "Date",
    y = "Prevalence (%)",
    title = "Differences in Mental Health Access by Race",
    caption = "Data was collected from August 19, 2020 to May 9, 2022")
```