---
title: "EDA_by_state"
author: "Caleigh Dwyer"
date: "2023-11-26"
output: html_document
---

```{r}
library(tidyverse)

library(sf)

library(maps)

library(plotly)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

##visualizations of US map require install of sf and maps packages
```

import tidy dataset:

```{r}
##note: should we export tidy dataset to our directory on github so we can delete this code from any EDA?

rawdata =
  read_csv("data/mental_health.csv")

tidydata = 
  rawdata |> 
  janitor::clean_names() |>
  mutate(indicator = str_replace(indicator, ", Last 4 Weeks", "")) |> 
  select(-phase, -quartile_range, -suppression_flag) |> 
  mutate(group = str_replace(group, "By ", "")) |>
  drop_na(value) |>
  select(start_date = time_period_start_date, end_date = time_period_end_date, everything())

tidydata =
  tidydata |>
  mutate(week_number = match(time_period_label, unique(time_period_label))) |>
  mutate(
    start_dates = mdy(pull(tidydata, start_date)),
    end_dates = mdy(pull(tidydata, end_date))) |> 
  mutate(tpl = time_period_label) |>  
  mutate(year = as.numeric(str_extract(tpl, "\\d{4}")))|> 
  separate(col = time_period_label, into = c("week_label", "years"), sep = ", ", remove = FALSE, extra = "merge") |>
  select(indicator, year, start_dates, end_dates, week_number, week_label = time_period_label, state, group, subgroup, value, low_ci, high_ci, confidence_interval)
```



## EDA

This EDA will analyze mental health care access trends by state.

I will first create a dataframe that only includes state-level observations.

```{r}
state_df =
  tidydata |> 
  filter(group == "State")
```

### Indicator 1: Took prescription

Let's start with the mental health indicator "Took Prescription Medication for Mental Health"

The following chunk shows the top five states with the highest mean percentage of respondents who reported taking prescription medication for mental health from 2020-2022.

```{r}

state_meds_top5 =
  state_df |> 
  filter(indicator == "Took Prescription Medication for Mental Health") |> 
  group_by(state) |> 
  summarize(mean_value = mean(value)) |> 
  arrange(desc(mean_value)) |> 
  top_n(5) |> 
  knitr::kable()

```


Let's visualize a heat map of this indicator's prevalence by US state.

First, we'll merge the filtered dataset with an existing US state map dataset to assist with mapping.

```{r}
state_meds =
  state_df |> 
  filter(indicator == "Took Prescription Medication for Mental Health") |> 
  group_by(state) |> 
  summarize(mean_value = mean(value)) |> 
  mutate(state = tolower(state))

usa = 
  st_as_sf(map("state", fill = TRUE, plot = FALSE))

us_meds =
  merge(usa, state_meds, by.x = "ID", by.y = "state", all.x = TRUE)
```


Next, we'll plot the data using ggplot

```{r}
heatmap_meds=
ggplot(us_meds)+
  geom_sf(aes(fill = mean_value), color = "white", size = 0.2)+
  scale_fill_gradient(low = "white", high = "red", na.value = "grey50", name = "Avg % taking prescription medications")+
  theme_minimal()+
  theme(legend.position = "right",
        plot.title.position = "plot")+
  labs(title = "Average % used prescription medication for mental health 2020-2022, by state")
```


We can also look at which states have had the greatest change in medication use. This shows that not only does WV have the highest prevalence of medication use for mental health during this period, but it also has had the highest level of change in this indicator between 2020-2022.

```{r}
state_meds_change =
  state_df |> 
  filter(indicator == "Took Prescription Medication for Mental Health", year != "2021") |> 
  group_by(state, year) |> 
  summarize(mean_value = mean(value)) |> 
  arrange(state, year) |> 
  mutate(val_change = c(NA, diff(mean_value)))

state_meds_change |> 
  filter(!is.na(val_change)) |> 
  ungroup() |> 
  select(-year) |> 
  slice_max(order_by = val_change, n=5) |> 
  knitr::kable()
  
```


Let's take a look at the trend for this indicator over time. 

```{r}
state_meds_trend =
state_df |>
  filter(indicator == "Took Prescription Medication for Mental Health") |> 
  group_by(state, year) |>
  summarize(mean_value = mean(value)) |> 
  ggplot(aes(x = year, y = mean_value, color = state)) +
  geom_line()
  
print(state_meds_trend)


```


```{r, fig.keep="none"}
state_meds_plotly=
state_df |>
  filter(indicator == "Took Prescription Medication for Mental Health") |> 
  group_by(state, year) |>
  summarize(mean_value = mean(value)) |> 
plot_ly(x = ~year, y = ~mean_value, color = ~state, type = "scatter", mode = "lines+markers") |>
  layout(
    title = "% used prescription medication for mental health by state & year",
    xaxis = list(title = "Year"),
    yaxis = list(title = "Average % used prescription medication for mental health"))

print(state_meds_plotly)
```

