---
title: "Statistical Analysis"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: "hide"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include = FALSE}
library(tidyverse)
library(modelr)
library(mgcv)
library(MASS) # for stepAIC
library(dplyr)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

tidydata = 
  read.csv("data/tidydata.csv")
```

This page shows the bivariate analysis and linear modeling we undertook as additional analyses to the Exploratory Data Analysis (EDA).

## Bivariate Analysis

We will conduct bivariate analysis to see if there is a statistical relationship between each demographic characteristic and two of our indicators of mental health access:

1. Took prescription medication and/or received counseling or therapy for mental health
2. Needed counseling or therapy for mental health but did not get it.

In other words, we will test whether group means for each demographic characteristic are statistically different for each indicator of mental health access. P-values less than 0.05 will be considered as statistically significant.

Since gender is binary, we will run an independent samples t-test. The remaining demographic characteristics (age group, education level, race/ethnicity, and sexual orientation) all have three or more categories, so we will run one-way ANOVA tests within them.

Even though these tests assume normality, our sample size (50,000-90,000) is large. The Central Limit Theorem allows these statistical tests to be used even when the population distribution is not normal because the distribution of the sample means will tend to be normal in such large sample sizes.

### Gender

We will first conduct independent samples t-tests to see if there is a statistical relationship between gender and each of the two aforementioned indicators of mental health access.

#### Took prescription medication and/or received counseling or therapy for mental health

```{r}
gender_received =
  tidydata |>
  filter(
    indicator == "Took Prescription Medication for Mental Health And/Or Received Counseling or Therapy",
    state == "United States",
    group == "Sex",
    subgroup %in% c("Male", "Female"))

ttest_gender_received = t.test(value ~ subgroup, data = gender_received) |>
  broom::tidy() |>
  knitr::kable()

ttest_gender_received
```

The first independent samples t-test above shows statistically significant differences in mean prescription medication and/or counseling or therapy for mental health rates between men and women. More specifically, women had statistically higher mean rates of taking prescription medication for mental health and/or receiving counseling or therapy than men from 2020 to 2022 in the United States.

#### Needed counseling or therapy but did not get it

```{r}
gender_needed =
  tidydata |>
  filter(
    indicator == "Needed Counseling or Therapy But Did Not Get It",
    state == "United States",
    group == "Sex",
    subgroup %in% c("Male", "Female"))

ttest_gender_needed = t.test(value ~ subgroup, data = gender_needed) |>
  broom::tidy() |>
  knitr::kable()

ttest_gender_needed
```

The second independent samples t-test above shows statistically significant differences in mean rates of needing counseling or therapy for mental health but not getting it between men and women. More specifically, women had statistically higher mean rates of needing counseling or therapy for mental health but not getting it than men from 2020 to 2022 in the United States.

### Age Group

We will conduct ANOVA to see if there is a statistical relationship between age group and each of the two aforementioned indicators of mental health access.

#### Took prescription medication and/or received counseling or therapy for mental health

```{r}
age_received =
  tidydata |>
  filter(
    indicator == "Took Prescription Medication for Mental Health And/Or Received Counseling or Therapy",
    state == "United States",
    group == "Age") |>
  group_by(indicator)

anova_age_received = aov(value ~ subgroup, data = age_received)

anova_age_received |>
  broom::tidy() |>
  knitr::kable()
```

The first ANOVA above shows statistically significant differences in mean prescription medication and/or counseling or therapy for mental health rates between age groups. In other words, at least one age group had statistically different mean rates of taking prescription medication for mental health and/or receiving counseling or therapy from 2020 to 2022 in the United States.

#### Needed counseling or therapy but did not get it

```{r}
age_needed =
  tidydata |>
  filter(
    indicator == "Needed Counseling or Therapy But Did Not Get It",
    state == "United States",
    group == "Age") |>
  group_by(subgroup)

anova_age_needed = aov(value ~ subgroup, data = age_needed)

anova_age_needed |>
  broom::tidy() |>
  knitr::kable()
```

The second ANOVA above shows statistically significant differences in mean rates of needing counseling or therapy for mental health but not getting it between age groups. In other words, at least one age group had statistically different mean rates of needing counseling or therapy for mental health but not getting it from 2020 to 2022 in the United States.

### Education Level

We will conduct ANOVA to see if there is a statistical relationship between education level and each of the two aforementioned indicators of mental health access.

#### Took prescription medication and/or received counseling or therapy for mental health

```{r}
education_received =
  tidydata |>
  filter(
    indicator == "Took Prescription Medication for Mental Health And/Or Received Counseling or Therapy",
    state == "United States",
    group == "Education") |>
  group_by(subgroup)

anova_education_received = aov(value ~ subgroup, data = education_received)

anova_education_received |>
  broom::tidy() |>
  knitr::kable()
```

The first ANOVA above shows statistically significant differences in mean prescription medication and/or counseling or therapy for mental health rates between education levels. In other words, at least one education level had statistically different mean rates of taking prescription medication for mental health and/or receiving counseling or therapy from 2020 to 2022 in the United States.

#### Needed counseling or therapy but did not get it

```{r}
education_needed =
  tidydata |>
  filter(
    indicator == "Needed Counseling or Therapy But Did Not Get It",
    state == "United States",
    group == "Education") |>
  group_by(subgroup)

anova_education_needed = aov(value ~ subgroup, data = education_needed)

anova_education_needed |>
  broom::tidy() |>
  knitr::kable()
```

The second ANOVA above shows statistically significant differences in mean rates of needing counseling or therapy for mental health but not getting it between education levels. In other words, at least one education level had statistically different mean rates of needing counseling or therapy for mental health but not getting it from 2020 to 2022 in the United States.

### Race/Ethnicity

We will conduct ANOVA to see if there is a statistical relationship between race/ethnicity and each of the two aforementioned indicators of mental health access.

#### Took prescription medication and/or received counseling or therapy for mental health

```{r}
race_received =
  tidydata |>
  filter(
    indicator == "Took Prescription Medication for Mental Health And/Or Received Counseling or Therapy",
    state == "United States",
    group == "Race/Hispanic ethnicity") |>
  group_by(subgroup)

anova_race_received = aov(value ~ subgroup, data = race_received)

anova_race_received |>
  broom::tidy() |>
  knitr::kable()
```

The first ANOVA above shows statistically significant differences in mean prescription medication and/or counseling or therapy for mental health rates between racial/ethnic groups. In other words, at least one racial/ethnic group had statistically different mean rates of taking prescription medication for mental health and/or receiving counseling or therapy from 2020 to 2022 in the United States.

#### Needed counseling or therapy but did not get it

```{r}
race_needed =
  tidydata |>
  filter(
    indicator == "Needed Counseling or Therapy But Did Not Get It",
    state == "United States",
    group == "Race/Hispanic ethnicity") |>
  group_by(subgroup)

anova_race_needed = aov(value ~ subgroup, data = race_needed)

anova_race_needed |>
  broom::tidy() |>
  knitr::kable()
```

The second ANOVA above shows statistically significant differences in mean rates of needing counseling or therapy for mental health but not getting it between racial/ethnic groups. In other words, at least one racial/ethnic group had statistically different mean rates of needing counseling or therapy for mental health but not getting it from 2020 to 2022 in the United States.

### Sexual Orientation

We will conduct ANOVA to see if there is a statistical relationship between sexual orientation and each of the two aforementioned indicators of mental health access.

#### Took prescription medication and/or received counseling or therapy for mental health

```{r}
sexuality_received =
  tidydata |>
  filter(
    indicator == "Took Prescription Medication for Mental Health And/Or Received Counseling or Therapy",
    state == "United States",
    group == "Sexual orientation",
    week_number %in% c(24:33)) |>
  group_by(subgroup)

anova_sexuality_received = aov(value ~ subgroup, data = sexuality_received)

anova_sexuality_received |>
  broom::tidy() |>
  knitr::kable()
```

The first ANOVA above shows statistically significant differences in mean prescription medication and/or counseling or therapy for mental health rates between sexual orientation groups. In other words, at least one sexual orientation group had statistically different mean rates of taking prescription medication for mental health and/or receiving counseling or therapy from 2020 to 2022 in the United States.

#### Needed counseling or therapy but did not get it

```{r}
sexuality_needed =
  tidydata |>
  filter(
    indicator == "Needed Counseling or Therapy But Did Not Get It",
    state == "United States",
    group == "Sexual orientation",
    week_number %in% c(24:33)) |>
  group_by(subgroup)

anova_sexuality_needed = aov(value ~ subgroup, data = sexuality_needed)

anova_sexuality_needed |>
  broom::tidy() |>
  knitr::kable()
```

The second ANOVA above shows statistically significant differences in mean rates of needing counseling or therapy for mental health but not getting it between sexual orientation groups. In other words, at least one sexual orientation group had statistically different mean rates of needing counseling or therapy for mental health but not getting it from 2020 to 2022 in the United States.

## Linear Modeling

First, we will import the tidy data set.

```{r}
library(tidyverse)

tidydata = 
  read.csv("data/tidydata.csv")
```

Then, we will separate the tidy data set into 4 subgroups: one for each indicator of mental health access.

```{r}
Indicator1=tidydata |> filter(indicator=="Took Prescription Medication for Mental Health")

Indicator2=tidydata |> filter(indicator=="Received Counseling or Therapy")

Indicator3=tidydata |> filter(indicator=="Took Prescription Medication for Mental Health And/Or Received Counseling or Therapy")

Indicator4=tidydata |> filter(indicator=="Needed Counseling or Therapy But Did Not Get It")
```

Next, we will apply a linear model for each indicator using stepwise regression.

```{r}
fit <- lm(value ~ group + subgroup + year + state, data=Indicator1)
stepwise_fit <- stepAIC(fit, direction="both")

fit <- lm(value ~ group + subgroup + year + state, data=Indicator2)
stepwise_fit <- stepAIC(fit, direction="both")

fit <- lm(value ~ group + subgroup + year + state, data=Indicator3)
stepwise_fit <- stepAIC(fit, direction="both")

fit <- lm(value ~ group + subgroup + year + state, data=Indicator4)
stepwise_fit <- stepAIC(fit, direction="both")
```

AIC values are identical across the steps shown, which suggests that removal or addition of a variable did not improve the model's fit.

In an attempt to use a linear model, we used the stepwise function to see if adding the variable `subgroup` was useful in identifying the best linear model for our data set. Upon further analysis, we realized that the stepwise approached was ineffective given that all our predictors are under one variable: subgroup. The AIC values confirmed this as they were identical across each step.

We attempted to rearranged the data using the `pivot_wider` function to isolate each predictor to a distinct variable rather than having all of our predictors under the subgroup variable. However, we came across challenges. First, the data for each predictor has a different value for each week. Second, the number the predictors differed across weeks. Given that our data set is grouped data, we were unable to organize our data set in a way that allowed us to isolated predictors and apply a linear model with each demographic characteristic as a predictor.