---
title: "Statistical Analysis"
author: "Sarah Younes"
date: "2023-11-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message = FALSE}
library(tidyverse)

tidydata = 
  read.csv("data/tidydata.csv")
```

Focus on two indicators:

1. "Took Prescription Medication for Mental Health And/Or Received Counseling or Therapy"
2. "Needed Counseling or Therapy But Did Not Get It"

## Independent samples t-tests: gender

### Took prescription medication for mental health and/or received counseling or therapy

```{r}
gender_received =
  tidydata |>
  filter(
    indicator == "Took Prescription Medication for Mental Health And/Or Received Counseling or Therapy",
    state == "United States",
    group == "Sex",
    subgroup %in% c("Male", "Female")
  ) |>
  select(week_number, value, subgroup) |>
  pivot_wider(
    names_from = "subgroup",
    values_from = "value") |>
  janitor::clean_names()

gender_received |>
  summarise(
    t_statistic = t.test(male, female)$statistic,
    p_value = t.test(male, female)$p.value)
```

### Needed counseling or therapy but did not get it

```{r}
gender_needed =
  tidydata |>
  filter(
    indicator == "Needed Counseling or Therapy But Did Not Get It",
    state == "United States",
    group == "Sex",
    subgroup %in% c("Male", "Female")
  ) |>
  select(week_number, value, subgroup) |>
  pivot_wider(
    names_from = "subgroup",
    values_from = "value") |>
  janitor::clean_names()

gender_needed |>
  summarise(
    t_statistic = t.test(male, female)$statistic,
    p_value = t.test(male, female)$p.value)
```

## ANOVA: sexual orientation

```{r}
sexuality_received =
  tidydata |>
  filter(
    indicator == "Took Prescription Medication for Mental Health And/Or Received Counseling or Therapy",
    state == "United States",
    group == "Sexual orientation",
    week_number %in% c(24:33)) |>
  select(week_number, value, subgroup) |>
  pivot_wider(
    names_from = "subgroup",
    values_from = "value") |>
  janitor::clean_names()
```

```{r}
sexuality_needed =
  tidydata |>
  filter(
    indicator == "Needed Counseling or Therapy But Did Not Get It",
    state == "United States",
    group == "Sexual orientation",
    week_number %in% c(24:33)) |>
  select(week_number, value, subgroup) |>
  pivot_wider(
    names_from = "subgroup",
    values_from = "value") |>
  janitor::clean_names()
```



#Check normality 
```{r}
ggplot(tidydata, aes(sample = log(value))) +
  stat_qq() +
  stat_qq_line()

 hist(log(tidydata$value), breaks = "FD", main = "Histogram of Data", xlab = "Data Values")
```



